\texttt{escheR}: Optimize Spatial Visualization Following Gestalt Principles

### 2023-03-17 Manuscript/manuscript.tex
add plots, add captions, set up cross-references

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context\cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations\cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE \cite{hinton_2002} and UMAP \cite{becht_2019}) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side, one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations\cite{sriworarat_2023} or 3-D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

 To optimize \textit{in situ} visualization, we introduce gestalt principles to the design of spatial maps. Gestalt principles\cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Following gestalt principles, we add additional visual dimensions to spatial maps, leveraging different aesthetics, such as color, fill, and symbols, to simultaneously display disparate variables. Specifically, we apply figure-ground articulation to display two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles, inspired by watercolor effect\cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground articulation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles\cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

\section*{Use Case 1: Spatially Differential Colocalization in DLPFC}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex (DLPFC) \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), are identified to colocalize via the cell-cell communication analysis. In addition, data suggest Layer 6 is the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we apply \texttt{escheR} to create a multi-dimensional \textit{in-situ} spatial map (Figure \label{fig:visual}) that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are ploted in two side-by-side figures, our proposed visualization enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

\section*{Use Case 2: Multi-dimensional UMAP}
The application of the proposed framework doesn't limit to in-situ visualizations of spatially resolved data. It can be broadly applied to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and t-SNE. To demonstrate, we apply the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, \texttt{schex} proposed to apply hexagonal binning strategy to unbiasedly display the gene expression.\cite{freytag_2020} In addition, color-coded convex hulls were used to annotate different clusters of cells (Figure \label{fig:visual}). Nevertheless, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we integrate the proposed visualization to \texttt{schex} by replacing convex hulls with color-coded hexagons boundaries  (Figure \label{fig:visual}) to avoid possible membership confusion. The visual improvement is easily implemented without any modification of \texttt{schex} thanks to the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

\section*{Data availability}
The spatial transcriptomics dataset was obtained from \texttt{spatialLIBD} (\url{http://research.libd.org/spatialLIBD/}). The UMAP example follows the tutorial from \texttt{schex} (\url{https://www.bioconductor.org/packages/release/bioc/vignettes/schex/inst/doc/using_schex.html}). The code that generates these figures are deposited at \fixme{github manuscript repo}.

\clearpage
% \vspace{-0.1in}
\begin{wrapfigure}[13]{r}{0.8\textwidth}
\vspace{-0.3in}
\begin{center}
\includegraphics[width=0.78\textwidth]{figure/new_visiual.png}
\end{center}
\vspace{-0.35in}
\caption{\footnotesize \textbf{Optimized visualization interpreting co-localization in the spatial context}. (\textbf{A}) Current visualization of co-localization requires two separate plots, one to display spatial domain information and one to display co-localization, difficulty to associate the two sources of information for interpretation. (\textbf{B}) The proposed visualization is optimized to display both spatial domain and co-localization information in one plot.}
\label{fig:visual} 
\end{wrapfigure}

>>>>

The recent technology breakthrough in spatial profiling enables characterizing the diversity of molecules, e.g. DNA, RNA, and proteins, in the spatial context\cite{moffitt_2022}, generating novel spatially-resolved data that requires highly tailored \textit{in situ} visualizations\cite{dries_2021, Lewis_2021, odonoghue_2021}. Current \textit{in situ} visualizations, as opposed to embedding visualizations (e.g. t-SNE \cite{hinton_2002} and UMAP \cite{becht_2019}) that project information to some mathematical space, visualize molecular information in the original cellular location, creating spatial maps that color-code a single variable (e.g., molecule expression, cell type, or spatial domain). This approach lacks the ability to simultaneously exhibit multiple variables, possibly from disparate data domains (such as expression domain and spatial domain) or modalities (such as transcriptomics and proteomics), creating cognitive gaps to interpret findings regarding their (micro-) environment. For example, when studying differential expression in different spatial domains (commonly defined by tissue architecture, e.g. cancerous regions, cortical layers), two spatial maps are displayed side-by-side (Figure \textbf{\ref{fig:visual}A-B}), one for gene expression and one for spatial domains, creating significant challenges to associate these two information together. While software-based interactive visualizations\cite{sriworarat_2023} or 3D visualizations have the potential to mitigate this challenge, they are infeasible for scientific communications in static media, such as prints. Developing a static spatial visualization that enables the simultaneous display of multiple dimensions of information is crucial for spatially-resolved data research.

@@

 To optimize \textit{in situ} visualization, we introduce gestalt principles to the design of spatial maps. Gestalt principles\cite{todorovic_2008, palmer_1999} refer to a set of rules describing how humans perceive and interpret visual information and are commonly applied in art and designs. Following gestalt principles, we add additional visual dimensions to spatial maps, leveraging different aesthetics, such as color, fill, and symbols, to simultaneously display disparate variables. Specifically, we apply figure-ground articulation to display two variables: one variable (e.g. expression) can be plotted as color-filled circles, serving as the \textit{figure}; one variable (e.g. spatial domains) can be plotted as the backgrounds of the circles, creating a \textit{ground} for the figure. For adjacent circles with limited space between them to display the background color, we use an economic implementation, colored outlines for these circles (Figure \textbf{\ref{fig:visual}C}), inspired by watercolor effect\cite{pinna_1987, pinna_2001}. Watercolor effect describes the phenomenon in visual perception that surface color arises from thin boundaries and hence is applied here to perceive the background color in tight space. Overall, the figure-ground articulation creates two isolated layers in visual perception to display the two variables while maintaining the relative spatial relationship serving as a reference between the two. In addition, other fundamental principles\cite{todorovic_2008}, such as proximity, similarity, continuity, and closure, incentivize the brain to group elements and dimensions in the visualization, guaranteeing an integrative perception of the complex multi-dimensional spatial map.

@@

\section*{Use Case 1: Spatially Differential Colocalization in DLPFC}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex (DLPFC) \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), are identified to colocalize via the cell-cell communication analysis. In addition, data suggest Layer 6 is the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we apply \texttt{escheR} to create a multi-dimensional \textit{in-situ} spatial map (Figure \textbf{\ref{fig:visual}A-B}) that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are ploted in two side-by-side figures, our proposed visualization (Figure \textbf{\ref{fig:visual}D}) enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

\section*{Use Case 2: Multi-dimensional UMAP}
The application of the proposed framework doesn't limit to in-situ visualizations of spatially resolved data. It can be broadly applied to data mapped to any 2-dimensional coordinate system to simultaneously display multiple variables. Such systems include euclidean space (including spatial coordinate as a special case) and data-driven embedding space, for example, UMAP and t-SNE. To demonstrate, we apply the proposed visualization to address the challenge of simultaneously displaying cluster membership and gene expression in a single-cell UMAP plot. To address the overplotting problem, \texttt{schex} proposed to apply hexagonal binning strategy to unbiasedly display the gene expression.\cite{freytag_2020} In addition, color-coded convex hulls were used to annotate different clusters of cells (Figure \textbf{\ref{fig:visual}E}). Nevertheless, the convex hulls create substantial overlapping areas, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. To improve the interpretability of the visualization, we integrate the proposed visualization to \texttt{schex} by replacing convex hulls with color-coded hexagons boundaries  (Figure \textbf{\ref{fig:visual}F}) to avoid possible membership confusion. The visual improvement is easily implemented without any modification of \texttt{schex} thanks to the Grammar of Graphics \cite{wilkinson_2012} standard.

@@

\section*{Data availability}
The spatial transcriptomics dataset was obtained from \texttt{spatialLIBD} (\url{http://research.libd.org/spatialLIBD/}). The UMAP example follows the tutorial from \texttt{schex} (\url{https://www.bioconductor.org/packages/release/bioc/vignettes/schex/inst/doc/using_schex.html}). The code that generates these figures is deposited at \url{https://github.com/boyiguo1/Manuscript_escheR}.

\clearpage

\begin{figure}[13]
\begin{center}
\includegraphics[width=0.78\textwidth]{Manuscript/figure/fig_grid.png}
\end{center}
\caption{\footnotesize \textbf{\texttt{escheR} improves spatial visualizations following the gestalt principles}. (\textbf{A-B}) The traditional visualization displays the cortex layers (spatial domains) (\textbf{A}) and the colocalization status (\textbf{B}) in a side-by-side figure, creating challenges to connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \texttt{in-situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability. (\textbf{E}) \texttt{schex} creates overlapped convex hulls, creating confusion when interpreting cluster memberships of hexagons in the overlapping areas. (\textbf{F}) \texttt{escheR} plots hexagon-specific membership to improve interpretability.}
\label{fig:visual} 
\end{figure}
